Staged Abstract Interpreters

### 2019-03-23 paper/21_bg_interp.tex
[paper] monadic sec

In Scala, thanks for its nature that combines functional programming and
object-oriented programming, we may use traits and objects to implement monads.
A trait poses monadic operations as abstract methods that objects extending it
should implement. An object encapsulates the data and implements monadic methods
that operate on these data. The monadic @bind@ operation is usually called
@flatMap@ in Scala. For example, the following trait @Monad@ declares the
@return@ and @flatMap@ operation:

\begin{lstlisting}
  trait Monad[F[_]] {
    def return[A](a: => A): F[A]
    def flatMap[A,B](fa: F[A])(f: A => F[B]): F[B]
  }
\end{lstlisting}

An object may encapsulate the data of type @F[A]@ and only provide
@flatMap(f: A => F[B]): F[B]@. This allows us to utilize the for-comprehensions in
Scala, which provide special syntactic support for moandic @flatMap@.
For example, a list object of element type @A@ is a monad and it implements the
methods @flatMap[B](f: A => List[B]): List[B]@ and @map[B](f: A => B): List[B]@.
To compute the Cartesian product of two lists of numbers, we may directly call
@flatMap@ and pass an anonymous function that internally calls the monadic
@return@ operation (the @List@ constructor in this case):

\begin{lstlisting}
  val xs = List(1, 2); val ys = List(4, 5)
  xs.flatMap(x => ys.flatMap(y => List((x, y)))) // List((1,4), (1,5), (2,4), (2,5))
\end{lstlisting}

Or, we can use for-comprehensions that computes the same value:

\begin{lstlisting}
  for {x <- xs; y <- ys} yield (x, y) // List((1,4), (1,5), (2,4), (2,5))
\end{lstlisting}

The Scala compiler will translate the above for-comprehension expression into
an equivalent one using @flatMap@ and @map@ ~\cite{scala_spec}:

@@

In the rest of the paper, we will use for-comprehension syntax for monadic
operations, the implementation of monads and monad transformers essentially
borrows the ground-truth from Haskell.

>>>>

In Scala, thanks for its nature that combines functional programming and
object-oriented programming, we may use traits and objects to implement monads.
For example, the trait @Monad@ (Figure \ref{fig:monad}) declares the @pure@ and @flatMap@
operation. The trait itself takes a higher-kinded type @M[_]: * -> *@ as
argument, which takes a type and returns a type. The method @pure@ promotes
values of type @A@ to values of type @M[A]@ \footnote{We elect to use
\texttt{pure} as the name, since \texttt{return} is a keyword in Scala and
\texttt{unit} is a built-in function in LMS.}. The monadic @bind@ operation is
usually called @flatMap@ in Scala, which takes a monad-encapsulated value of
type @M[A]@, a function of @A => M[B]@ and returns values of type @M[B]@.

\begin{figure}[h!]
  \centering
  \begin{subfigure}[b]{0.55\textwidth}
    \begin{lstlisting}
  trait Monad[M[_]] {                                  
    def pure[A](a: A): M[A]                            
    def flatMap[A,B](fa: M[A])(f: A => M[B]): M[B]     
  }                                                    
    \end{lstlisting}
    \caption{trait \texttt{Monad}} \label{fig:monad}
  \end{subfigure}
  ~
  \begin{subfigure}[b]{0.4\textwidth}
    \begin{lstlisting}
trait MonadOps[M[_], A] {
  def flatMap[B](f: A => M[B]): M[B]
  def map[B](f: A => B): M[B]
}
    \end{lstlisting}
    \caption{trait \texttt{MonadOps}} \label{fig:monadops}
  \end{subfigure}
\end{figure}

Similar to Haskell's @do@-notation, Scala provides special syntactic support for
monadic operations through @for@-comprehension.
For example, a object of @List[A]@, where @A@ is the element type. Then to
compute the the Cartesian product of two lists of numbers, we can use Scala's
@for@-comprehension syntax.

\begin{lstlisting}
  val xs = List(1, 2); val ys = List(4, 5)
  for {x <- xs; y <- ys} yield (x, y) // List((1,4), (1,5), (2,4), (2,5))
\end{lstlisting}

The Scala compiler will translate the above @for@-comprehension expression into
an equivalent one using @flatMap@ and @map@ ~\cite{scala_spec}. The last binding
in the @for@-comprehension is translated into a @map@, where the expression of
@yield@ becomes the body expression of that @map@ application. The foregoing
bindings are all translated into calls of @flatMap@.

@@

Note that here the monadic object @List[_]@ encapsulates the data internally.
Therefore it only exposes simplified version of @flatMap@ and @map@, where the
monad @M[A]@ is not introduced as a function argument. The trait @MonadOps@
(Figure \ref{fig:monadops}) defines the simplified version of monadic
operations. The conversion between @Monad@ and @MonadOps@ can be done by using
implicit design pattern.
In the rest of the paper, we use Scala's for-comprehension syntax and monad
transformers such as @ReaderT@, @StateT@, and @ListT@ to write our interpreters.
The implementation of monads and monad transformers essentially borrows the
ground-truth from Haskell.
